---
title: "BST 260 Final Project"
output: github_document
date: "2022-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis focuses on the sexual health of American adolescents using data from the National Longitudinal Survey of Youth which followed a cohort of adolescents born 1980-1984 beginning in 1997 to 2019. The survey is sponsored by the U.S. Bureau of Labor Statistics and includes 8,894 respondents. At the time of the first interview in 1997, respondents were aged 12 to 18, and are now 34 to 40 years of age. 51% of the original sample was male and 49% was female. 51.9% of the sample was Non-Black/non-Hispanic, 26% was Black non-Hispanic, 21.2% of the sample was Hispanic or Latino, and 0.9% reported being mixed race. This analysis uses data from the year 2002. 

Using this data I sought to: 
* Determine if there is an association between sex and ever being diagnosed with an STI. 
* Determine if there is an association between use of Condom during last sexual intercourse and ever being diagnosed with an STI.
* Determine the effect of knowledge about best form of birth control to prevent STI transmission on the likelihood of reporting ever being diagnosed with an STI. 

The variables I use in this analysis are:

* Sex: Defined as a binary male or female.
* Race: Defined as Non-Black/non-Hispanic, Black non-Hispanic, Hispanic or Latino, or mixed race. 
* Condom: If the respondent or their partner used a condom at the time of last sex. 
* STI: If the respondent has ever been diagnosed with an STI other than HIV/AIDS. 

I begin by preparing the data: 

```{r}
#Load packages
library(tidyverse)
library(lmtest)

# Load the data
data <- read_csv("C:/Users/alisc/OneDrive - Harvard University/Harvard/Fall 2022/BST 260/BST260_FinalProject/default.csv", show_col_types = FALSE) 

# Rename columns to more clearly reflect the variables
colnames(data) <- c("ID", "Sex", "BD_M", "BD_Y", "Sample_Type", "Race", "Condom", "Other", "HIV", "STI", "K_preg", "K_STI")
View(data)

# Create dummy variables for demographics 
data$female <- ifelse(data$Sex == 2, 1, 0)
data$Black <- ifelse(data$Race == 2, 1, 0)
data$Hispanic <- ifelse(data$Race == 1, 1, 0)

# Create dummy variable for use of modern birth control method at time of last sexual intercourse 
data$bc <- ifelse(data$Condom == 1 | data$Other == 2 | data$Other == 4 | data$Other == 7 | data$Other == 11 | data$Other == 6 | data$Other == 8 | data$Other == 10, 1, 0)

# Create dummy variable for correct knowledge about best best control method to prevent STI transmission out of the following options: withdrawal, condom, birth control pill
data$K <- ifelse(data$K_STI == 2, 1, 0)
```

Some exploratory data analysis: 

The proportion of respondents who reported using a condom at the time of last sexual intercourse is 0.125. The proportion of respondents who reported using any form of modern birth control at the time of last sexual intercourse is 0.146. The proportion of respondents who were able to correctly identify that a condom is the most effective form of birth control to prevent STI transmission was 0.164. 

```{r}
##### DEAL WITH THIS (below)
##Is there a 
### Show the count data
```

* Create a variable to deal with non-responses for K *

```{r}
prop.table(table(data$Condom))
prop.table(table(data$bc))
prop.table(table(data$K))
```

My approach to answering my research question will begin with a simple test of association to determine if there is statistically significant evidence of an association between sex and reporting ever being diagnosed with an STI. The data is count data and all cells in the 2x2 table are greater than 5, so I will use a Pearson's Chi-Square test. To determine if there is an association between using a condom at last sexual intercourse and reporting ever being diagnosed with an STI, I will use a Fisher's Exact Test because the data is binomial and not all cells in the 2x2 table are greater than 5. Then, I will conduct a logistic regression to determine the effect of having correct knowledge about what form of birth control is most effective for preventing STI transmission on reporting ever being diagnosed with an STI. I will estimate the effect using a logit and a probit model and then conduct a likelihood ratio test to determine which is a better estimation. I also will control for variables such as sex and race. 

**Results**

Tests of Association

```{r}
# Remove non-response variables
data$Condom[data$Condom < 0] <- NA
data$STI[data$STI < 0] <- NA
data$HIV[data$HIV < 0] <- NA

# Is there an association between sex and reporting ever being diagnosed with an STI?
data$Sex <- factor(data$Sex,
                   levels = c(1, 2),
                   labels = c("Male", "Female"))
table(data$STI, data$Sex)
chisq.test(data$STI, data$Sex, correct=FALSE)
```

This test indicates that there is a statistically significant association between sex and reporting ever being diagnosed with an STI (p = 0.0002968). 

```{r}
# Is there an association between use of Condom during last sex and reporting an STI?
data$Condom <- factor(data$Condom,
                      levels = c(0, 1),
                      labels = c("No Condom", "Yes Condom"))

data$STI <- factor(data$STI,
                   levels = c(0, 1),
                   labels = c("No STI", "Yes STI"))

table(data$Condom, data$STI)
fisher.test(data$Condom, data$STI)
```
This test indicates that based on these data, there is not a statistically significant association between the respondent or their partner using a condom at the time of last sexual intercourse and reporting ever being diagnosed with an STI. 

Logistic Regression 
```{r}
# Logit model
m1 <- glm(STI ~ K + female + Black + Hispanic,  family = binomial(link = "logit"), data = data)
summary(m1)

# Probit model 
m2 <- glm(STI ~ K + female + Black + Hispanic,  family = binomial(link = "probit"), data = data)
summary(m2)

# Likelihood Ratio Test
lrtest(m1,m2)
```
This analysis indicates that the probit model specified above is the better fit of the two. The probit model predicts that being able to correctly identify that condoms are the best form of birth control to prevent STI transmission, compared to withdrawal and the birth control pill, reduces the likelihood of reporting ever being diagnosed with an STI by 0.283. However, these findings are not statistically significant. Being female, however, did have a statistically significant 
