---
title: "BST 260 Final Project"
output: github_document
date: "2022-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis focuses on the sexual health of American adolescents using data from the National Longitudinal Survey of Youth which followed a cohort of adolescents born 1980-1984 beginning in 1997 to 2019. The survey is sponsored by the U.S. Bureau of Labor Statistics and includes 8,894 respondents. At the time of the first interview in 1997, respondents were aged 12 to 18, and are now 34 to 40 years of age. 51% of the original sample was male and 49% was female. 51.9% of the sample was Non-Black/non-Hispanic, 26% was Black non-Hispanic, 21.2% of the sample was Hispanic or Latino, and 0.9% reported being mixed race. This analysis uses data from the year 2002. 

Using this data I sought to: 

* Determine if there is an association between sex and ever being diagnosed with a sexually transmitted infection (STI). 
* Determine if there is an association between use of condom during last sexual intercourse and ever being diagnosed with a STI.
* Determine the effect of being female on the likelihood of reporting ever being diagnosed with a STI. 

The variables I use in this analysis are:

* Sex: Defined as a binary male or female.
* Race: Defined as Non-Black/non-Hispanic, Black non-Hispanic, Hispanic or Latino, or mixed race. 
* Condom: If the respondent or their partner used a condom at the time of last sexual intercourse. 
* STI: If the respondent has ever been diagnosed with a STI other than HIV/AIDS. 

I begin by preparing the data: 

```{r, echo = FALSE}
#Load packages
library(tidyverse)
library(lmtest)

# Load the data
data <- read_csv("C:/Users/alisc/OneDrive - Harvard University/Harvard/Fall 2022/BST 260/BST260_FinalProject/default.csv", show_col_types = FALSE) 

# Rename columns to more clearly reflect the variables
colnames(data) <- c("ID", "Sex", "BD_M", "BD_Y", "Sample_Type", "Race", "Condom", "Other", "HIV", "STI", "K_preg", "K_STI")

head(data)

# Create dummy variables for demographics 
data$female <- ifelse(data$Sex == 2, 1, 0)
data$Black <- ifelse(data$Race == 2, 1, 0)
data$Hispanic <- ifelse(data$Race == 1, 1, 0)

# Remove respondents that skipped the question about condom use at time of last sexual intercourse and about if they have ever been diagnosed with an STI
data <- data[data$Condom != -4 & data$Condom != -5, ] 
data <- data[data$STI != -4 & data$STI != -5, ] 

# Create dummy variable for use of modern birth control method at time of last sexual intercourse 
data$bc <- ifelse(data$Condom == 1 | data$Other == 2 | data$Other == 4 | data$Other == 7 | data$Other == 11 | data$Other == 6 | data$Other == 8 | data$Other == 10, 1, 0)
```

```{r, echo = FALSE}
data$HIV[data$HIV < 0] <- 0
data$HIV <- as.numeric(data$HIV)
sum(data$HIV)
```

I was hoping to investigate both STIs and HIV/AIDS diagnoses, but in this dataset, there was only 1 HIV diagnosis, so the analysis would not be useful or valid. 

```{r, echo = FALSE}
prop.table(table(data$Condom))
```

The proportion of respondents who reported that they or their partner used a condom at the time of last sexual intercourse is 0.824.

```{r, echo = FALSE}
prop.table(table(data$bc))
```

The proportion of respondents who reported using any form of modern birth control at the time of last sexual intercourse is 0.959. I define modern birth control as: condom, foam, jelly, cream, sponge, or suppository, diaphragm, IUD, morning after pill, birth control pill, depo-provera or injectable, or norplant. I do not include withdrawal or rhythm (safe time). 

My approach to answering my research question will begin with a simple test of association to determine if there is statistically significant evidence of an association between sex and reporting ever being diagnosed with an STI. The data is binomial data and not all cells in the 2x2 table are greater than 5, so I will use a Fisher's Exact Test. To determine if there is an association between using a condom at last sexual intercourse and reporting ever being diagnosed with a STI, I will use a Fisher's Exact Test as well. Then, I will conduct a binomial regression to determine the effect of having correct knowledge about what form of birth control is most effective for preventing STI transmission on reporting ever being diagnosed with a STI. I will estimate the effect using a logit and a probit model while controlling for race and then conduct a likelihood ratio test to determine which is a better estimation. 

**Results**

Tests of Association

Is there an association between sex and reporting ever being diagnosed with a STI?

```{r, echo = FALSE}
# Is there an association between sex and reporting ever being diagnosed with a STI?
data$STI <- factor(data$STI,
                   levels = c(0, 1),
                   labels = c("No STI", "Yes STI"))

data$Sex <- factor(data$Sex,
                   levels = c(1, 2),
                   labels = c("Male", "Female"))
table(data$STI, data$Sex)
fisher.test(data$STI, data$Sex)
```

This test indicates that based on these data, there is a statistically significant association between sex and reporting ever being diagnosed with a STI (p = 0.01514). 

Is there an association between use of condom during last sexual intercourse and reporting a STI?
```{r, echo = FALSE}
# Is there an association between use of Condom during last sexual intercourse and reporting a STI?
data$Condom <- factor(data$Condom,
                      levels = c(0, 1),
                      labels = c("No Condom", "Yes Condom"))

table(data$Condom, data$STI)
fisher.test(data$Condom, data$STI)
```

This test indicates that based on these data, there is not a statistically significant association between the respondent or their partner using a condom at the time of last sexual intercourse and reporting ever being diagnosed with a STI. 

Binomial Regression 

What is the effect of being female on the likelihood of reporting ever being diagnosed with a STI? Beginning with a logit model: 
```{r, echo = FALSE}
# Logit model
data$STI <- as.numeric(data$STI)
data$STI <- ifelse(data$STI == 2, 1, 0)

m1 <- glm(STI ~ female + Black + Hispanic,  family = binomial(link = "logit"), data = data)
summary(m1)

ggplot(data, aes(x=female, y=STI)) + 
  geom_point() +
  stat_smooth(method="glm", se=FALSE, method.args = list(family=binomial(link = "logit")))
```

This analysis indicates that being female significantly reduces the likelihood of reporting ever being diagnosed with a STI by 2.21 points (p <= 0.01). There is no significant effect of being Black or Hispanic on the likelihood of reporting ever being diagnosed with a STI. 

How does the estimate change when using a probit model?

```{r, echo = FALSE}
# Probit model

m2 <- glm(STI ~ female + Black + Hispanic,  family = binomial(link = "probit"), data = data)
summary(m2)

ggplot(data, aes(x=female, y=STI)) + 
  geom_point(alpha=.5) +
  stat_smooth(method="glm", se=FALSE, method.args = list(family=binomial(link = "probit")))
```

This probit model predicts that being female significantly reduces the likelihood of reporting ever being diagnosed with a STI by 0.987 points (p <= 0.01). There is no significant effect of being Black or Hispanic on the likelihood of reporting ever being diagnosed with a STI. 

Is the logit or probit model a better fit?

```{r, echo = FALSE}
# Likelihood Ratio Test
lrtest(m1,m2)
```

This analysis indicates that the logit model specified above is the better fit of the two.

**Summary**

These analyses answered the following questions: 

* Is there an association between sex and ever being diagnosed with a STI?
* Is there an association between use of condom during last sexual intercourse and ever being diagnosed with a STI?
* What is the effect of being female on the likelihood of reporting ever being diagnosed with a STI? 

To answer these questions, I used tests of association and binomial regression. I found that there is an association between sex and ever being diagnosed with a STI (p = 0.01514), but there is not evidence of an association between use of a condom during last sexual intercourse and ever being diagnosed with a STI. A logistic regression predicted that being female significantly reduces the likelihood of reporting ever being diagnosed with a STI by 2.21 points (p <= 0.01) while a probit model predicted that being female significantly reduces the likelihood of reporting ever being diagnosed with a STI by 0.987 points (p <= 0.01). Neither model predicted any significant effect of being Black or Hispanic on the likelihood of reporting ever being diagnosed with a STI. A likelihood ratio test identified the logit model as the better fit for the data. 

This analysis was successful in answering my questions based on the data available. If I had more time I would be interested in finding data that had greater internal validity. The quality of the data is limited in that the researchers asked about biological sex but not gender. Furthermore, sex is a nuanced concept that many people interpret differently (e.g. what counts as "sex"), and by asking about practices at last sexual intercourse, the researchers many not be capturing the nature of non-heterosexual sex. Furthermore, STI transmission can occur during sexual acts that are not intercourse. So, as a preliminary analysis, this is interesting, but the data collection methods could be more robust. I also worry about desirability bias in these data as respondents may have altered their answer to seem more in accordance with health recommendations. 

**References**

Bureau of Labor Statistics, U.S. Department of Labor. National Longitudinal Survey of Youth 1997 cohort, 1997-2017 (rounds 1-18). Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH: 2019.

**Appendix: All code for this report**
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

